---
title: "p8105_hw3_yz4993"
author: "Yixin Zheng"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(readxl)
library(dplyr)
library(knitr)
library(lubridate)
library(viridis)
library(patchwork)
library(ggridges)
library(hexbin)
library(janitor)
library(rnoaa)
library(p8105.datasets)

knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

# Problem 1

```{r 1 load data, include=FALSE}
data("ny_noaa")

# check structure
str(ny_noaa)
```

This dataset contains `r nrow(ny_noaa)` rows and `r ncol(ny_noaa)` columns. Variables include weather station id, date of observation,  (tenths of mm), snowfall (mm), snow depth (mm), and min and max temperature (tenths of degrees C).

Below we clean the data, creating separate variables for year, month, and day and converting `tmax` and `tmin` to numeric. We find that 0 is the most commonly observed value for snowfall. This is because most days of the year, it does not snow at all in NY. The second most commonly observed value is `NA`, indicating missingness. Other common values are 13, 25, and 51, suggesting that snowfall is originally recorded in fractions of an inch and converted to mm.

```{r 1.1 data cleaning + snowfall value}
ny_noaa %>% 
  count(snow) %>%
  arrange(desc(n))

ny_noaa = 
  ny_noaa %>% 
  separate(date, into = c("year", "month", "day"), convert = TRUE) %>% 
  mutate(
    tmax = as.numeric(tmax),
    tmin = as.numeric(tmin))
```


Below is a two-panel plot showing the average max temperature in January and in July in each station across years. As expected, the mean temperature in January is much lower than the mean temperature in July for all stations and across all years. All stations appear to follow similar trends of temperature peaks and valleys within a month across the years, i.e. when one station has a high monthly mean temperature for a given year, most other stations also have a high monthly mean temperature for that year. We do see one uncharacteristically cold station in July of 1987 or 1988, as well as a few other less drastic outliers.

```{r 1.2 plot avg tmax}
ny_noaa %>% 
  group_by(id, year, month) %>% 
  filter(month %in% c(1, 7)) %>% 
  summarize(mean_tmax = mean(tmax, na.rm = TRUE, color = id)) %>% 
  ggplot(aes(x = year, y = mean_tmax, group = id)) + geom_point() + geom_path() +
  facet_grid(~month) +
  labs(title = "Mean monthly temperature for each station across years for January and July")
```

Below we show a two-panel plot including (i) a hex plot of `tmax` vs `tmin` for the full dataset; and (ii) a ridge plot showing the distribution of snowfall values (in mm) greater than 0 and less than 100 separately by year. 

From the hex plot we see that while there is some variability, the majority of the data cluster tightly in the center of the distribution. In relatively rare cases, it seems that `tmax` is less than `tmin`, which raises questions about data recording and quality.

From the ridge plot, we see a multimodal density of snowfall within a given year. Most stations see between 0 and  35 mm of snow in a year. Then there is a another group of stations that see about 45 mm of snow, and another group that sees nearly 80 mm. It is likely this multimodality stems from the conversion of measurements in one system (fractions of an inch) to another (using the metric system), which was also noted in the table of common values. 

```{r 1.3 plot tmax vs tmin and snowfall}
hex = 
  ny_noaa %>% 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_hex()

ridge = 
  ny_noaa %>% 
  filter(snow < 100, snow > 0) %>%
  ggplot(aes(x = snow, y = as.factor(year))) + 
  geom_density_ridges()

hex + ridge
```
# Problem 2
```{r 2.1 Load, tidy, merge,, include=FALSE}
# Load and clean the datasets
demographic_data <- read_csv("~/Desktop/Data Science/HW/HW3/p8105_hw3_yz4993/data/nhanes_covar.csv", skip = 4) %>%
  clean_names() %>%
  filter(age >= 21, !is.na(sex), !is.na(education)) %>%
  mutate(
    # convert sex to a factor
    sex = factor(sex, levels = c(1, 2), labels = c("Male", "Female")),
    
    # convert education to a factor
    education = factor(education, 
                       levels = c(1, 2, 3), 
                       labels = c("Less than high school", "High school equivalent", "More than high school"),
                       ordered = TRUE)
  )

accelerometer_data <- read_csv("~/Desktop/Data Science/HW/HW3/p8105_hw3_yz4993/data/nhanes_accel.csv") %>%
  clean_names() %>%
  pivot_longer(cols = starts_with("min"), 
               names_to = "minute", 
               values_to = "activity", 
               names_prefix = "min") %>%
  mutate(minute = as.numeric(minute))

# merge datasets
merged_data <- inner_join(demographic_data, accelerometer_data, by = "seqn")
```


Below, the table shows the number of men and women across three education categories: "Less than high school," "High school equivalent," and "More than high school." Both gender are roughly equally distributed in the "Less than high school" and "More than high school" categories, while in "High school equivalent" category there are more men than women.
```{r 2.2 reader-friendly table and visualization}
gender_education_table <- merged_data %>%
  group_by(sex, education) %>%
  summarize(count = n_distinct(seqn)) %>%
  pivot_wider(
    names_from = education,
    values_from = count,
    names_prefix = "Education: "
  ) %>%
  knitr::kable()

gender_education_table

merged_data %>%
  filter(!is.na(education)) %>%
  ggplot(aes(x = age, fill = sex)) +
  geom_density(alpha = 0.5) +
  facet_grid(. ~ education) +
  scale_fill_manual(values = c("blue", "orange"), name = "Sex", labels = c("Male", "Female")) +
  labs(
    title = "Age Distribution by Education Category and Gender",
    x = "Age",
    y = "Density",
    caption = "Data from NHANES Study"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```
The age distribution plot below the table indicates that both male (in blue) and female (in orange) follow a similar age pattern in the "More than high school" category, with female peaking around age 30 (y-axis represents density), and male peaking around 23-24. Both genders show a decrease in density as age increases within this category. In the other two categories, male and female exhibit different age patterns.
In the "Less than high school" category, male have the highest density around ages 45-47, with the distribution being left-skewed. Despite a trough around age 57, there are secondary peaks between ages 60-80. For female in this category, the highest peak is around age 80, with the distribution also being left-skewed, showing an increase in density as age increases.
In the "High school equivalent" category, female again show a left-skewed distribution with a peak around age 80, while male have a right-skewed distribution with a peak around age 21.

```{r 2.3 aggregate and plot}
# aggregate across minutes to create a total activity variable for each participant
total_activity_data <- merged_data %>%
  group_by(seqn, sex, age, education) %>%
  summarize(total_activity = sum(activity, na.rm = TRUE)) %>%
  ungroup()

total_activity_data %>%
  filter(!is.na(education)) %>%
  ggplot(aes(x = age, y = total_activity, color = sex)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "loess", se = FALSE) +
  facet_grid(. ~ education) +
  scale_color_manual(values = c("blue", "orange"), name = "Sex", labels = c("Male", "Female")) +
  labs(
    title = "Total Activity vs Age by Education Category and Gender",
    x = "Age",
    y = "Total Activity",
    caption = "Data from NHANES Study"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```
The 3-panel plot above shows total daily activity plotted against age, with separate panels for different education levels. Generally, activity levels decrease with age across all education categories for both genders, but there are notable exceptions. In both the "High school equivalent" and "More than high school" categories, females exhibit slightly higher activity levels compared to males. In the "Less than high school" category, females have higher activity levels than males before age 50, but lower levels after age 50. Additionally, in the "High school equivalent" category, females show significantly higher activity levels than males around age 40.


```{r 2.4 3-panel plot}
# convert minute to hours
activity_timecourse_data <- merged_data %>%
  mutate(hour = minute / 60) %>%  # 
  group_by(hour, education, sex) %>%
  summarize(avg_activity = mean(activity, na.rm = TRUE)) %>%
  ungroup()

activity_timecourse_data %>%
  filter(!is.na(education)) %>%
  ggplot(aes(x = hour, y = avg_activity, color = sex)) +
  geom_line() +  # Add the activity line
  geom_smooth(method = "loess", se = FALSE) +  # smooth trend lines
  facet_grid(. ~ education) +  #  by education
  scale_color_manual(values = c("blue", "orange"), name = "Sex", labels = c("Male", "Female")) +
  labs(
    title = "24-Hour Activity Time Course by Education Level and Sex with Smooth Trends",
    x = "Hour of the Day",
    y = "Average Activity",
    caption = "Data from NHANES Study"
  ) +
  scale_x_continuous(breaks = seq(0, 24, by = 2), labels = seq(0, 24, by = 2)) +  # set x-axis breaks at every 2 hours
  theme_minimal() +
  theme(legend.position = "bottom")
```
The plot above also reflects similar observations to the previous one, with males and females having comparable average activity levels by the hour in the "Less than high school" category, while females exhibit noticeble higher average activity levels than males in the other two categories. It is notable that, across all education levels, both men and women follow similar hourly activity patterns—activity slowly increase from 00:00 to 10:00,  peaks around 10:00- 13:00, then a steady decline after.

